On the Effects of Varying Spatial Resolutions of 3DEP Data on Flood Inundation Extents Produced from Height Above Nearest Drainage

### 2022-10-04 manuscripts/owp_3dep_2022/owp_3dep.tex
added to intro

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}
%
%% Additional packages
\usepackage[printonlyused]{acronym}
%\usepackage{hyperref}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\journal{Advances in Water Resources}
%
\begin{document}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% FRONT MATTER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frontmatter}
%
%% Title, authors and addresses
%
%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%% \fntext[label3]{}

@@

\title{Investigating the Effect of Varying the Spatial Resolutions of LiDAR Derived Elevations on Flood Inundation Maps Produced from Height Above Nearest Drainage}

@@

\affiliation[nwc]{
             organization={National Water Center, Office of Water Prediction, National Oceanic and Atmospheric Administration},
             addressline={205 Hackberry Ln}, 
             city={Tuscaloosa},
             postcode={35401}, 
             state={VA},
             country={USA}
            }
%
\affiliation[uf]{
             organization={Center for Remote Sensing, Agricultural and Biological Engineering, University of Florida},
             addressline={1741 Museum Rd}, 
             city={Gainesville},
             postcode={32603}, 
             state={FL},
             country={USA}
            }
%
\affiliation[uh]{
             organization={Civil and Environmental Engineering, University of Houston},
             addressline={4226 Martin Luther King Boulevard}, 
             city={Houston},
             postcode={77204}, 
             state={TX},
             country={USA}
            }
%
\affiliation[usgs]{
             organization={Hydrologic Applied Innovations Lab, New York Water Science Center, United States Geological Survey},
             addressline={425 Jordan Rd}, 
             city={Troy},
             postcode={12180}, 
             state={NY},
             country={USA}
            }
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{abstract}

\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Graphical abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Research highlights
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{highlights}
\item Continental-scale flood inundation mapping requires simplifying assumptions
\item Utilizing LiDAR derived data from the USGS 3DEP program enhances the skill on flood inundation maps produced from \ac{HAND}.
\item Varying the spatial resolution provides no skill enhancement when evaluated at large scales across three study sites and benchmark datasets.
\end{highlights}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% KEYWORDS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% MAX SIX WORDS %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{keyword}
Hydrology \sep
Height Above Nearest Drainage \sep
Flood Inundation Mapping \sep
Digital Elevation Maps \sep
Forecasting
%
%% PACS codes here, in the form: \PACS code \sep code
\PACS 92.40.-t \sep 92.40.Qk \sep *92.40.Q- \sep *92.40.qp \sep 89.60.Gg \sep 92.60.Wc \sep 91.10.Jf \sep 84.40.Xb
%
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%
\end{keyword}
%
\end{frontmatter}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% MAIN TEXT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% prints line numbers
\linenumbers
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}
%
% Flush acronym usage
\acresetall 
%
%%%% Motivates and introduces flooding %%%%
Floods are among the most frequent, damaging, and deadly of natural disasters \citep{doocy2013human,stromberg2007natural,kahn2005death}. 
The frequency and intensity of flood events as well as the exposure of people and property to them have been increasing in recent times driven by secular changes in climate, infrastructure, and demographics \citep{berz2000flood,mallakpour2015changing,downton2005reanalysis,kunkel1999temporal,pielke2000precipitation,corringham2019effect}. 
Unfortunately, these trends are expected to continue placing additional pressure on hydrological extremes \citep{kahn2005death,tabari2020climate,milly2002increasing,wing2018estimates}.
Floods impact mortality and morbidity through drowning or physical trauma at the individual health scale, while increasing the risk of infectious disease at the public health level \citep{jonkman2005global,beinin2012medical,alajo2006cholera,french1983mortality}.
Flooding disrupts systems providing human needs such as transportation routes, supply chains, water delivery, waste management, communications, and energy grids \citep{wijkman2021natural}.
These impacts disproportionately affect certain demographics such as the socioeconomically-disadvantaged, youth, and elderly who are more likely to live in vulnerable areas with less access to educational resources, \acp{EWS}, and the capacity or resources to evacuate impacted areas \citep{kahn2005death,smiley2022social,stromberg2007natural,jonkman2005global,tellman2020using,tellman2021satellite}.
These inequitable impacts further entrench poverty and inequalities \citep{stallings1988conflict,birkmann2010extreme}.
In political terms, severe disasters, including floods, can reduce social order, strain governance systems, collapse social safety nets, increase the risk of social conflict \citep{drury1998disasters,xu2016natural,zahran2009natural}.
These dire consequences motivate adaption and mitigation efforts such as \acp{EWS}, protective infrastructure (e.g. storage, defenses, drainage, infiltration), public awareness and education, and zoning regulations \citep{tumbare2000mitigating,tauhid2018mitigating,charlesworth201115}.

%%%% Motivates and introduces NWM %%%%
Due to the growing consequences and risks presented by increasing flood impacts, \acp{EWS}, or forecasting systems, can help understand future conditions and provide intelligence to furnish adequate warnings to protect life, prevent damages, and enhance resilience \citep{stromberg2007natural,cools2016lessons,unisdr2015making,baudoin2014early,golnaraghi2012overview,unep2012early,liu2018review}.
The early warning of flood disasters at national scales requires the use of continental-scale, forecast hydrology models and modeling frameworks that span intranational political boundaries.
The applications of these models extend beyond \acp{EWS} to provide historical trends for applications in infrastructure planning, public planning, insurance underwriting, and more.
The \ac{OWP}, an office of the \ac{NOAA} along with partners at the \ac{NCAR}, developed such a continental-scale model known as the \ac{US} \ac{NWM} \citep{salas2018towards,gochis2021wrf,cosgrove2019evolution,cohen2018featured,noaa2016national,water2022nwm}.
The \ac{NWM} is based on a configuration of the \ac{WRF-Hydro} model that accounts for land surface processes as well as overland and channel routing \citep{gochis2021wrf,salas2018towards,cosgrove2019evolution}.
Operationally, the \ac{NWM} produces streamflow analysis and forecasts at multiple time horizons depending on location which include the \ac{CONUS}, \ac{PR}, and \ac{HI} \citep{cosgrove2019evolution,noaa2016national,water2022nwm}.
The \ac{NWM} routes streamflow across the \ac{NWM} \ac{V2.1} stream network, based on the \ac{NHDPlusV2} network, is comprised of more than 5.5 million \acp{km} of lines discretized into more than 2.8 million forecast points \citep{aristizabal2022reducing}.
The \ac{NWM} \ac{V2.1} stream network belongs to the \ac{NWM} ``hydrofabric'' defined as a catalog of geospatial layers relevant to hydrology modeling including stream network lines, catchments, reservoirs, and more \citep{water2022nwm,cosgrove2019evolution}.
While streamflow is an important variable for engineering and scientific applications of fluvial flooding, flood inundation stages, extents and depths are much more tangible variables to the stakeholders flood events directly impact.

%%%% Introduces and motivates HAND %%%%
The \acp{SWE}, a system of two hyperbolic partial differential equations, formally govern the flow of fluvial surface water by conserving both mass (first equation) and momentum (second equation) and can be expressed in both the \ac{1D} (Saint Venant Equations) and \ac{2D} forms.
Solving this system in full \ac{2D} form requires numerical methods that can be very cost prohibitive and numerically unstable in an operational setting across continental-scales at high spatial discretizations (10 \ac{m} or higher).
This use case motivates the implementation of an inundation proxy, also known as a zero-physics or a simplified conceptual model, that is agnostic to the \acp{SWE} while still computing accurate fluvial inundation extents and depths for this problem \citep{teng2015rapid}.
\ac{HAND} detrends elevations within \acp{DEM} to compute drainage potentials by normalizing elevations to the nearest, relevant drainage line instead of datums that represent mean sea level \citep{renno2008hand,nobre2011height,nobre2016hand}.
\ac{HAND} as a terrain index has been used extensively for \ac{FIM} purposes from both modeled or observed stream flows and stages \citep{nobre2016hand,afshari2018comparison,garousi2019terrain,johnson2019integrated,zheng2018geoflood,zheng2018river,zhang2018comparative,teng2015rapid,li2022accounting,li2020evaluation}, as well as for assisting the remote sensing detection of fluvial inundation \citep{aristizabal2020high,shastry2019using,aristizabal2021mapping,huang2017comparison,twele2016sentinel}.
\ac{HAND} operates as an inundation proxy by thresholding the relative elevation (or \ac{HAND}) values with a singular river stage value for each catchment corresponding to the drainage area of a given river reach \citep{nobre2016hand,garousi2019terrain,johnson2019integrated,zheng2018geoflood,teng2015rapid,li2020evaluation,liu2016cybergis,maidment2017conceptual,liu2018cybergis,liu2020height,liu2018review}.
When used to generate inundation extents and depths from streamflow, reach-averaged \ac{SRC} sample geometric variables along an entire reach and normalize using the length of the reach to create stage-discharge relationships \citep{zheng2018river,aristizabal2022reducing,godbout2019error}.
These relationships depend on a friction parameter, Manning's n, and used to convert streamflows to stages for eventual \ac{2D} mapping with \ac{HAND}.
Numerous investigations have validated the use of \ac{HAND} for flood mapping applications as a suitable alternative to more sophisticated physics-based techniques for large scale and high resolution use cases \citep{johnson2019integrated,li2020evaluation,li2022comparative,aristizabal2022reducing,nobre2016hand,godbout2019error,afshari2018comparison,zhang2018comparative,teng2015rapid,teng2017flood,diehl2021improving}.

>>>>

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}
%
%% Additional packages
\usepackage[printonlyused]{acronym}
\usepackage{multirow}
\usepackage{comment}
\usepackage{url}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\journal{Advances in Water Resources}
%
\begin{document}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% FRONT MATTER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frontmatter}
%
%% Title, authors and addresses
%
%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%% \fntext[label3]{}

@@

\title{On the Effects of Varying Spatial Resolutions of 3DEP Data on Flood Inundation Extents Produced from Height Above Nearest Drainage}

@@

\affiliation[nwc]{
             organization={National Water Center, Office of Water Prediction, National Oceanic and Atmospheric Administration},
             addressline={205 Hackberry Ln}, 
             city={Tuscaloosa},
             postcode={35401}, 
             state={VA},
             country={USA}
            }
%
\affiliation[uf]{
             organization={Center for Remote Sensing, Agricultural and Biological Engineering, University of Florida},
             addressline={1741 Museum Rd}, 
             city={Gainesville},
             postcode={32603}, 
             state={FL},
             country={USA}
            }
%
\affiliation[uh]{
             organization={Civil and Environmental Engineering, University of Houston},
             addressline={4226 Martin Luther King Boulevard}, 
             city={Houston},
             postcode={77204}, 
             state={TX},
             country={USA}
            }
%
\affiliation[usgs]{
             organization={Hydrologic Applied Innovations Lab, New York Water Science Center, United States Geological Survey},
             addressline={425 Jordan Rd}, 
             city={Troy},
             postcode={12180}, 
             state={NY},
             country={USA}
            }
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{abstract}
\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Graphical abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Research highlights
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{highlights}
\item HAND approximates flood inundation for large scale, high resolution, and operational applications.
\item Utilizing LiDAR derived data from the USGS 3DEP program enhances the skill on flood inundation maps produced from HAND.
\item Varying the spatial resolution provides no skill enhancement when evaluated at large scales across three study sites and benchmark datasets.
\end{highlights}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% KEYWORDS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% MAX SIX WORDS %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{keyword}
%% PACS codes here, in the form: \PACS code \sep code
Hydrology, surface water, water resources, water supply, floods, lakes, rivers, streamflow, natural and man-made disasters, weather analysis and prediction, topography

\PACS
92.40.-t \sep % hydrology, cryosphere
92.40.Qk \sep % surface water, water resources, water supply, floods, lakes, rivers, streamflow,
*92.40.Q- \sep % surface water
*92.40.qp \sep % streamflow, floods
89.60.Gg \sep % disasters (natural and man-made),
92.60.Wc \sep % weather analysis and prediction
91.10.Jf      % topography (earth)
%
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%
\end{keyword}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{frontmatter}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% MAIN TEXT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% prints line numbers
\linenumbers
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}
%
% Flush acronym usage
\acresetall 
%
%%%% Motivates and introduces flooding %%%%
Floods are among the most frequent, damaging, and deadly of natural disasters \citep{doocy2013human,stromberg2007natural,kahn2005death}. 
The frequency and intensity of flood events as well as the exposure of people and property to them have been increasing in recent times driven by secular changes in climate, infrastructure, and demographics \citep{berz2000flood,mallakpour2015changing,downton2005reanalysis,kunkel1999temporal,pielke2000precipitation,corringham2019effect}. 
Unfortunately, these trends are expected to continue placing additional pressure on hydrological extremes \citep{kahn2005death,tabari2020climate,milly2002increasing,wing2018estimates}.
Floods impact mortality and morbidity through drowning or physical trauma at the individual health scale, while increasing the risk of infectious disease at the public health level \citep{jonkman2005global,beinin2012medical,alajo2006cholera,french1983mortality}.
Flooding disrupts systems providing human needs such as transportation routes, supply chains, water delivery, waste management, communications, and energy grids \citep{wijkman2021natural}.
These impacts disproportionately affect certain demographics such as the socioeconomically-disadvantaged, youth, and elderly who are more likely to live in vulnerable areas with less access to educational resources, \acp{EWS}, and the capacity or resources to evacuate impacted areas \citep{kahn2005death,smiley2022social,stromberg2007natural,jonkman2005global,tellman2020using,tellman2021satellite}.
These inequitable impacts further entrench poverty and inequalities \citep{stallings1988conflict,birkmann2010extreme}.
In political terms, severe disasters, including floods, can reduce social order, strain governance systems, collapse social safety nets, increase the risk of social conflict \citep{drury1998disasters,xu2016natural,zahran2009natural}.
These dire consequences motivate adaption and mitigation efforts such as \acp{EWS}, protective infrastructure (e.g. storage, defenses, drainage, infiltration), public awareness and education, and zoning regulations \citep{tumbare2000mitigating,tauhid2018mitigating,charlesworth201115}.

%%%% Motivates and introduces NWM %%%%
Due to the growing consequences and risks presented by increasing flood impacts, \acp{EWS}, or forecasting systems, can help understand future conditions and provide intelligence to furnish adequate warnings to protect life, prevent damages, and enhance resilience \citep{stromberg2007natural,cools2016lessons,unisdr2015making,baudoin2014early,golnaraghi2012overview,unep2012early,liu2018review}.
The early warning of flood disasters at national scales requires the use of continental-scale, forecast hydrology models and modeling frameworks that span intranational political boundaries.
The applications of these models extend beyond \acp{EWS} to provide historical trends for applications in infrastructure planning, public planning, insurance underwriting, and more.
The \ac{OWP}, an office of the \ac{NOAA} along with partners at the \ac{NCAR}, developed such a continental-scale model known as the \ac{US} \ac{NWM} \citep{salas2018towards,gochis2021wrf,cosgrove2019evolution,cohen2018featured,noaa2016national,water2022nwm}.
The \ac{NWM} is based on a configuration of the \ac{WRF-Hydro} model that accounts for land surface processes as well as overland and channel routing \citep{gochis2021wrf,salas2018towards,cosgrove2019evolution}.
Operationally, the \ac{NWM} produces streamflow analysis and forecasts at multiple time horizons depending on location which include the \ac{CONUS}, \ac{PR}, and \ac{HI} \citep{cosgrove2019evolution,noaa2016national,water2022nwm}.
The \ac{NWM} routes streamflow across the \ac{NWM} \ac{V2.1} stream network, based on the \ac{NHDPlusV2} network, is comprised of more than 5.5 million \acp{km} of lines discretized into more than 2.8 million forecast points \citep{aristizabal2022extending}.
The \ac{NWM} \ac{V2.1} stream network belongs to the \ac{NWM} ``hydrofabric'' defined as a catalog of geospatial layers relevant to hydrology modeling including stream network lines, catchments, reservoirs, and more \citep{water2022nwm,cosgrove2019evolution}.
While streamflow is an important variable for engineering and scientific applications of fluvial flooding, flood inundation stages, extents and depths are much more tangible variables to the stakeholders flood events directly impact.

%%%% Introduces and motivates HAND %%%%
The \ac{SWE}, a system of two hyperbolic partial differential equations, formally govern the flow of fluvial surface water by conserving both mass (first equation) and momentum (second equation) and can be expressed in both the \ac{1D} (Saint Venant Equations) and \ac{2D} forms.
Solving this system in full \ac{2D} form requires numerical methods that can be very cost prohibitive and numerically unstable in an operational setting across continental-scales at high spatial discretizations (10 \ac{m} or higher).
This use case motivates the implementation of an inundation proxy, also known as a zero-physics or a simplified conceptual model, that is agnostic to the \acp{SWE} while still computing accurate fluvial inundation extents and depths for this problem \citep{teng2015rapid,bates2000simple}.
\ac{HAND} detrends elevations within \acp{DEM} to compute drainage potentials by normalizing elevations to the nearest, relevant drainage line instead of datums that represent mean sea level \citep{renno2008hand,nobre2011height,nobre2016hand}.
\ac{HAND} as a terrain index has been used extensively for \ac{FIM} purposes from both modeled or observed stream flows and stages \citep{nobre2016hand,afshari2018comparison,garousi2019terrain,johnson2019integrated,zheng2018geoflood,zheng2018river,zhang2018comparative,teng2015rapid,li2022accounting,li2020evaluation}, as well as for assisting the remote sensing detection of fluvial inundation \citep{aristizabal2020high,shastry2019using,aristizabal2021mapping,huang2017comparison,twele2016sentinel}.
\ac{HAND} operates as an inundation proxy by thresholding the relative elevation (or \ac{HAND}) values with a singular river stage value for each catchment corresponding to the drainage area of a given river reach \citep{nobre2016hand,garousi2019terrain,johnson2019integrated,zheng2018geoflood,teng2015rapid,li2020evaluation,liu2016cybergis,maidment2017conceptual,liu2018cybergis,liu2020height,liu2018review}.
When used to generate inundation extents and depths from streamflow, reach-averaged \ac{SRC} sample geometric variables along an entire reach and normalize using the length of the reach to create stage-discharge relationships \citep{zheng2018river,aristizabal2022extending,godbout2019error}.
These relationships depend on a friction parameter, Manning's n, and used to convert streamflows to stages for eventual \ac{2D} mapping with \ac{HAND}.
Numerous investigations have validated the use of \ac{HAND} for flood mapping applications as a suitable alternative to more sophisticated physics-based techniques for large scale and high resolution use cases \citep{johnson2019integrated,li2020evaluation,li2022comparative,aristizabal2022extending,nobre2016hand,godbout2019error,afshari2018comparison,zhang2018comparative,teng2015rapid,teng2017flood,diehl2021improving,hocini2021performance,bates2003optimal}.

%%%% Motivates and introduces OWP FIM %%%%
Several prior and active large-scale \ac{HAND} implementations catered to operational \ac{EWS} applications including the \ac{NFIE} \citep{maidment2017conceptual,liu2016cybergis,liu2018cybergis}, GeoFlood \citep{zheng2018geoflood,hocini2021performance,d2022identification,carruthers2021assessment,zheng2022application}, and \ac{PyGFT}, i.e. \ac{GIS} \citep{petrochenkov2020pygft,verdin2016software}.
The \ac{NFIE} was a broad, inter-institutional, and pioneering effort to apply HAND to the initial versions of the \ac{NWM} which leveraged 1/3 arc-second (10 \ac{m}) seamless elevation data available at the time \citep{maidment2017conceptual,liu2016cybergis,liu2018cybergis} from the \ac{USGS}'s \ac{NED} \citep{gesch2002national,gesch2007digital}.
\citet{zheng2018geoflood} applied HAND for operational applications with 1/27 arc-second (1 \ac{m}) elevation data with a novel least cost, geodesic based stream delineation method \citep{passalacqua2010geometric,passalacqua2012automatic,zheng2018geoflood,zheng2019automatic,carruthers2021assessment,d2022identification,zheng2022application}.
For applications with the \ac{NWM}, an advanced version of \ac{HAND} coupled with the use of \acp{SRC}, known as \ac{OWP} \ac{FIM}, converts \ac{NWM} analysis, reanalysis, and forecast streamflows to river stages and fluvial inundation depths and extents on an operational basis to \ac{CONUS} while extending the modeling domain to \ac{PR} and \ac{HI} \citep{aristizabal2022extending,inundationMapping2022}.
\ac{OWP} \ac{FIM} utilizes some of the latest datasets including the \ac{NHDPlusHR} \citep{moore2019user}, \ac{NLD} \citep{engineers2016national}, and the \ac{NWM} \ac{V2.1} hydrofabric \citep{water2022nwm,noaa2016national,nwm2022hydrofabric,gochis2021wrf}.
These datasets enforce hydrologically relevant features such as levees and the general location of stream lines to facilitate conflation with the forecast stream network \citep{aristizabal2022extending,inundationMapping2022}.
Additionally, \ac{OWP} \ac{FIM} advanced a fundamental limitation of \ac{HAND} that limits sourcing fluvial inundation only from the nearest, relevant drainage line \citep{mcgehee2016modified,aristizabal2022extending,zhang2018comparative,li2022comparative,zheng2018geoflood,zheng2018river,nobre2016hand}.
Stream lines of higher Horton-Strahler stream order that could contribute inundation to a given area have no way of extending beyond catchment lines which creates artificial bottlenecks in inundation extents especially along junctions of high order rivers with their lower flow tributaries \citep{aristizabal2022extending,mcgehee2016modified}.
To resolve this limitation, \ac{OWP} \ac{FIM} disaggregates the \ac{NWM} \ac{V2.1} stream network into segments of effective unit stream order called \acp{LP} \citep{aristizabal2022extending}.
In terms of terrain data, \ac{OWP} \ac{FIM} utilizes the 10 \ac{m} \ac{DEM} from the \ac{NHDPlusHR} elevation dataset which is the elevation basis, derived in batches from \ac{3DEP}, for additional hydrography products within the \ac{NHDPlusHR} \citep{aristizabal2022extending,moore2019user}.
The previous advances in \ac{OWP} \ac{FIM} stopped short of accounting for \ac{LiDAR} point elevation observations \citep{aristizabal2022extending} that are now nearing their first collection cycle to form a novel seamless, continental scale \ac{DEM} from \ac{3DEP} \citep{usgs2022status,usgs2022partnerships}.
